---
title: "Single-trait selection - Exercise"
author: "Leticia Aparecida de Castro Lara, Nelson Lubanga, and Gregor Gorjanc"
date: "2022-06-22"
---

# Introduction

In this exercise, you will simulate one generation of single-trait selection and evaluate the response to selection under different:

- Selection accuracy for one generation,
- Selection intensity for one generation, and
- EXTRA: Selection intensity over many generations.

For each step, we have given you instructions with an `AlphaSimR` template code to complete (replace ??? with an appropriate code)

# Base population

```{r}
# Clean the working environment
rm(list = ls())

# Set the default plot layout
par(mfrow = c(1, 1))

# Load AlphaSimR, simulate founder genomes, define a trait, and simulate a base population
library(AlphaSimR)
founderGenomes = runMacs(nInd = 100,
                         nChr = 10,
                         segSites = 100,
                         species = "MAIZE")
SP = SimParam$new(founderGenomes)
SP$addTraitA(nQtlPerChr = 100, mean = 10, var = 1)
basePop = newPop(founderGenomes)
```

# Accuracy of selection (one generation)

First, vary the accuracy of selection by varying the heritability of phenotypes used to identify the superior individuals. Assume heritability of 0.5 and select 50 individuals with the highest phenotype values.

```{r}
# Define parameters for the scenario with trait heritability of 0.5 and 50 selected
heritability = ???
nSelected = ???

# Phenotype base population
basePop = setPheno(basePop, h2 = ???)

# Select the best performing individuals according to their phenotype
basePopSelected = selectInd(pop = basePop,
                            nInd = ???,
                            use = "???")

# Create a new population and phenotype it
newPop = randCross(basePopSelected, nCrosses = nInd(basePop))
newPop = setPheno(newPop, h2 = ???)

# Evaluate observed response to selection between generations
# (as the difference between the mean of genetic values)
(deltaGObserved_h205 = ???(newPop) - ???(basePop))

# Quantify the intensity of selection
# (using the function selInt())
(selInt_h205 = selInt(p = ??? / ???))

# Quantify the accuracy of selection
# (as the correlation between genetic value and selection criterion)
(selAcc_h205 = cor(???(basePop), ???(basePop)))

# Quantify the standard deviation of genetic values
(sdGen_h205 = sqrt(???(basePop)))

# Predict expected response to selection between generations
(deltaGExpected_h205 = ??? * ??? * ???)
```

Now repeat the above code, but change trait heritability to 0.1 and save the results into objects `deltaGObserved_h201`, `selInt_h201`, `selAcc_h201`, `sdGen_h201`, and `deltaGExpected_h201`. Then repeat the above code by changing trait heritability to 0.9 and save the results into objects `deltaGObserved_h209`, `selInt_h209`, `selAcc_h209`, `sdGen_h209`, and `deltaGExpected_h209`.

Start with heritability of 0.1.

```{r}
# Define parameters for the scenario with trait heritability of 0.1 and 50 selected
???
```

Now with heritability of 0.9.

```{r}
# Define parameters for the scenario with trait heritability of 0.9 and 50 selected
???
```

# Intensity of selection (one generation)

Now repeat the above code, by changing trait heritability back to 0.5 and changing the number of selected individuals to 20 and save the results into objects `deltaGObserved_n20`, `selInt_n20`, `selAcc_n20`, `sdGen_n20`, and `deltaGExpected_n20`. Then change the number of selected individuals to 10 and save the results into objects `deltaGObserved_n10`, `selInt_10`, `selAcc_n10`, `sdGen_n10`, and `deltaGExpected_n10`.

Start with 20 selected individuals.

```{r}
# Define parameters for the scenario with trait heritability of 0.5 and 20 selected
???
```

Now with 10 selected individuals.

```{r}
# Define parameters for the scenario with trait heritability of 0.5 and 10 selected
???
```

Finally, combine all the results into a data.frame and discuss the observed and expected response to selection across the scenarios.

```{r}
results = data.frame(h2 = c(0.5, 0.1, 0.9, 0.5, 0.5),
                     nSelected = c(50, 50, 50, 20, 10),
                     deltaGObserved = c(deltaGObserved_h205, deltaGObserved_h201, deltaGObserved_h209, deltaGObserved_n20, deltaGObserved_n10),
                     deltaGExpected = c(deltaGExpected_h205, deltaGExpected_h201, deltaGExpected_h209, deltaGExpected_n20, deltaGExpected_n10),
                     selInt = c(selInt_h205, selInt_h201, selInt_h209, selInt_n20, selInt_n10),
                     selAcc = c(selAcc_h205, selAcc_h201, selAcc_h209, selAcc_n20, selAcc_n10),
                     sdGen = c(sdGen_h205, sdGen_h201, sdGen_h209, sdGen_n20, sdGen_n10))
print(results)
```

What do the results show you about changing these parameters?

# EXTRA: Intensity of selection (many generations)

Lastly, use the code below and evaluate the effect of the intensity of selection on response to selection over 50 generations in the following three steps:

  1) Run the simulation with 10 selected individuals and save results into objects `meanGAll_n10` and `varGAll_n10`.

  2) Repeat the simulation with 50 selected individuals and save results into objects `meanGAll_n50` and `varGAll_n50`.

  3) Plot trend in the mean and variance of genetic values over generations for both scenarios and discuss results.

Start with 10 selected individuals.

```{r}
# Set simulation parameters for the scenario with trait heritability of 0.5 and 10 selected
heritability = 0.5
nSelected = ???

# Phenotype base population
basePop = setPheno(basePop, h2 = heritability)

# Select the best performing individuals according to their phenotype
basePopSelected = selectInd(pop = basePop,
                            nInd = nSelected,
                            use = "pheno")

# Allocate vectors
nGenerations = ??? + 1 # +1 to store starting generation
meanGAll = numeric(nGenerations)
varGAll = numeric(nGenerations)

# Save the starting values
meanGAll[1] = meanG(basePop)
varGAll[1] = varG(basePop)

# To make the for loop below simpler we will make a copy of the object basePopSelected
newPopSelected = basePopSelected

# Selection over many generations
for (generation in 1:(nGenerations - 1)) {
  # Cross parents, phenotype progeny, and select new parents
  newPop = randCross(newPopSelected, nCrosses = nInd(basePop))
  newPop = setPheno(newPop, h2 = heritability)
  newPopSelected = selectInd(pop = newPop,
                             nInd = nSelected,
                             use = "pheno")
  # Save summaries
  meanGAll[1 + generation] = meanG(newPop)
  varGAll[1 + generation] = varG(newPop)
}

# Now save these outputs by copying the objects
meanGAll_n10 = meanGAll
varGAll_n10 = varGAll
```

Now with 50 selected individuals.

```{r}
# Set simulation parameters for the scenario with trait heritability of 0.5 and 50 selected
???
```

```{r}
par(mfrow = c(2, 1),
    mar = c(4, 4, 1, 1))

# Plot mean of genetic values over time
meanRanges = range(c(meanGAll_n10, meanGAll_n50))
plot(x = 1:nGenerations, y = meanGAll_n10, type = "l", col = "black", lwd = 3,
     xlab = "Generation", ylab = "Mean of genetic values", ylim = meanRanges)
lines(x = 1:nGenerations, y = meanGAll_n50, type = "l", col = "purple", lty = 2, lwd = 3)
legend(x = "topleft", legend = c(10, 50), title = "nSelected",
       lwd = 3, lty = c(1, 2), col = c("black", "purple"), bty = "n")

# Plot variance of genetic values over time
varRanges = range(c(???, ???))
plot(x = 1:nGenerations, y = ???, type = "l", col = "black", lwd = 3,
     xlab = "Generation", ylab = "Variance of genetic values", ylim = varRanges)
lines(x = 1:nGenerations, y = ???, type = "l", col = "purple", lty = 2, lwd = 3)
legend(x = "topright", legend = c(10, 50), title = "nSelected",
       lwd = 3, lty = c(1, 2), col = c("black", "purple"), bty = "n")
```

What do the results show you about changing these parameters?
